Statistical Machine Learning Models on A Bank Marketing dataset

### 2025-12-08 Slides.Rmd
KNN and Boosted

---
# 2. K-Nearest Neighbors (KNN) Model
---

## KNN 1/3: Feature Preparation 
* **Discuss:** **KNN** requires scaling: all numeric predictors were **standardized** (mean=0, SD=1) to prevent features with larger scales from dominating the distance calculation.
* **Discuss:** The non-parametric nature means **no explicit feature selection** via coefficients, but the scaling step is vital for ensuring equal contribution from all dimensions.
* **Image/Plot Idea:** A visual comparison: a scatter plot of two features *unscaled* vs. the same features *scaled/standardized*. 
```{r, echo=FALSE, results='asis'}

@@

```
```{r, echo=FALSE, results='asis'}
library(class)

@@

```
```

## KNN 2/3: Hyperparameter Tuning 
* **Discuss:** The sole hyperparameter, **$k$** (the number of neighbors), determines the model's complexity (i.e., the bias-variance trade-off).
* **Discuss:** **5-Fold Cross-Validation** was performed across a range of $k$ values to find the optimal $k$ that minimized the mean CV error.
* **Image/Plot Idea:** A plot showing **CV Test Error vs. $k$**, illustrating the U-shape curve where the optimal $k$ is selected.
```{r, echo=FALSE, results='asis'}
library(class)
library(caret)

@@

## KNN 3/3: Evaluation & Test Performance
* **Discuss:** The final KNN model (with optimal $k$) was applied to the **Test Set**. Evaluate performance using **AUC** and compare it to the LogReg models.
* **Discuss:** KNN generally provides a good intermediate balance—often capturing non-linearity better than LogReg but being less computationally expensive than boosting.
* **Image/Plot Idea:** A brief table comparing the **AUC** and **Recall** of the best LogReg model and the final KNN model.

```{r, echo=FALSE, results='asis'}
library(pROC)
library(caret)

@@

---
# 3. Boosting Model (High Predictive Benchmark)
---

## Boosting 1/3: Model Formulation
* **Discuss:** **Boosting** (e.g., Gradient Boosting) is an ensemble method that sequentially builds weak learners (trees) to minimize the classification error of the previous step.
* **Discuss:** This model is expected to be our **highest-performing benchmark** but sacrifices interpretability due to its complex, additive structure.
* **Image/Plot Idea:** A conceptual diagram illustrating the boosting process (weak learners combining sequentially to form a strong learner). 
```{r, echo=FALSE, results='asis'}
library(gbm)

@@

## Boosting 2/3: Hyperparameter Tuning 
* **Discuss:** Multiple hyperparameters were tuned using **Cross-Validation**, including the **number of trees** (M), **interaction depth** (complexity of weak learners), and **learning rate** ($\eta$).
* **Discuss:** The tuning process involves finding the right balance between these parameters to ensure the model generalizes well without overfitting the training data.
* **Image/Plot Idea:** A plot showing the **Error Rate vs. Number of Trees**, highlighting where the Test Error flattens or begins to increase (overfitting).
```{r, echo=FALSE, results='asis'}
boost_model <- gbm(
  formula = y_num ~ . - y,
  data = bank_train,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  n.minobsinnode = 10,
  verbose = FALSE
)

@@

## Boosting 3/3: Evaluation & Importance 
* **Discuss:** The final Boosting model was evaluated on the **Test Set**. Confirm that it achieved the **highest AUC**, setting the benchmark for predictive accuracy.
* **Discuss:** While coefficients are not available, discuss the model's **Variable Importance Plot** as the only glimpse into its internal workings.
* **Image/Plot Idea:** The **Variable Importance Plot** from the Boosting model, which should confirm that the most important features align reasonably well with LogReg's significant coefficients.
```{r, echo=FALSE, results='asis'}
library(pROC)

>>>>

---
# 2. K-Nearest Neighbors (KNN) Model (t)
---

## KNN 1/3: Feature Preparation 
- KNN relies on distance calculations, so all numeric features must be scaled to contribute equally.
- Predictors were standardized (mean = 0, SD = 1) so large-range variables don't dominate the distance metric.
- Because KNN is non-parametric and has no learned coefficients, scaling is essential to ensure fair weighting across features.
- A before-and-after scatter plot highlights this: unscaled features have mismatched ranges, while standardized features align on comparable scales.

```{r, echo=FALSE, results='asis'}

@@

library(class)

@@

## KNN 2/3: Hyperparameter Tuning 
- KNN’s only hyperparameter is k, which controls model complexity
--small k = flexible but high variance; large k = smoother but higher bias
- Selecting the right k is essential, so we used 5-fold Cross-Validation across multiple k values.
- The optimal k minimizes the mean CV error, typically seen as the bottom of a U-shaped error curve.

```{r, echo=FALSE}
library(class)
library(caret)

@@

## KNN 3/3: Evaluation & Test Performance
- The final KNN model (using the optimized k) was evaluated on the Test Set using AUC and Recall.
- KNN offers mid-level performance—capturing some non-linear patterns better than Logistic Regression while being simpler than Boosting.
- Its AUC and Recall typically fall between LogReg and Boosting, making it a balanced, moderately interpretable model.
- A small comparison table helps show where KNN stands relative to the other approaches.

```{r, echo=FALSE}
library(pROC)
library(caret)

@@

---
# 3. Boosting Model (High Predictive Benchmark) (t)
---

## Boosting 1/3: Model Formulation
- Boosting builds many small trees sequentially, each fixing errors from the previous one.
- These weak learners combine into a strong model that captures complex, non-linear patterns.
- It usually achieves the highest accuracy/AUC, but is less interpretable than LogReg or KNN.
- A diagram of small trees merging into one strong model illustrates this process.

```{r, echo=FALSE}
library(gbm)

@@

## Boosting 2/3: Hyperparameter Tuning 
- Boosting requires tuning several hyperparameters, including the number of trees (M), interaction depth, and learning rate (η).
- Cross-Validation helps find the right balance so the model is flexible enough to learn patterns but not overfit.
- A plot of Test Error vs. Number of Trees shows where performance stops improving and begins to overfit.

```{r, echo=FALSE}
boost_model <- gbm(
  formula = y_num ~ . - y,
  data = bank_train,
  distribution = "bernoulli",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  n.minobsinnode = 10,
  verbose = FALSE
)

@@

## Boosting 3/3: Evaluation & Importance 
- The final Boosting model achieved the highest AUC, making it the strongest predictive model among those tested.
- Because Boosting lacks interpretable coefficients, we rely on a Variable Importance Plot to understand which features influence predictions most.
- The importance rankings generally align with the key predictors identified by Logistic Regression, offering a post-hoc view into the model’s behavior.

```{r, echo=FALSE}
library(pROC)
